---
title: "Aggregate H3K9mes signals analysis around putative ZNF430 binding sites in THE1B"
author: Zheng Zuo
date: April, 2023
output:
  html_document:
    df_print: paged
    theme: lumen
    highlight: pygments
    toc: yes
    toc_float: yes
  pdf_document: default
---

## Extracting ZNF430 binding sites from all human THE1B elements

TECookbook package is needed here for extracting putative ZNF430 binding sites, which can be accessed from [here](https://github.com/zeropin/TECookbook)

```{r eval = FALSE}
require(dplyr)
require(GenomicRanges)
require(TECookbook)

ZNF430.THE1B.sites = TECookbook::liftOut("../Transposon Elements/hg38.fa.align",
                                         Repeat = "THE1B",
                                         start_pos = 197, end_pos = 222)

ZNF430.THE1B.sites = subset(ZNF430.THE1B.sites, seqnames %in% paste0("chr",c(1:23,"X","Y")))

ZNF430.full.sites  = subset(ZNF430.THE1B.sites, width == 26)

ZNF430.full.sites$Mismatch = TFCookbook::countMismatch(ZNF430.full.sites$Sequence,
                                                       "TCTTGCCTGCCGCCATGTAAGACGTG")

save("ZNF430.full.sites", file = "ZNF430.full.sites.RData")
```


## The sequence logo for all full-length putative ZNF430 binding sites
```{r fig.height=0.9, fig.width=5, message=FALSE, warning=FALSE}
require(dplyr)
require(GenomicRanges)
require(ggseqlogo)
load("ZNF430.full.sites.RData")
ZNF430.full.sites$Sequence %>%
  ggseqlogo::ggseqlogo()
```

### The distribution of mismatches to the consensus site within THE1B
```{r fig.height=1.2, fig.width=5, message=FALSE, warning=FALSE}
require(ggplot2)
ZNF430.full.sites %>% 
  as_tibble() %>%
  ggplot(aes(x = Mismatch))+
  geom_histogram(binwidth = 1)+
  theme_bw()
```


## Aggregate H3K9me3 signals around each class of sites in human placental tissues and hepatocytes

First of all, we need sort all putative ZNF430 binding sites into three classess(strong, intermediate, and weak) based on their mismatches to the consensus site within THE1B elements. The H3K9me3 dataset is publicly available from [ENCODE database](https://www.encodeproject.org/).

```{r message=FALSE, warning=FALSE, fig.height=3.3, fig.width=12}
require(soGGi)
ZNF430.Strong.sites       = subset(ZNF430.full.sites, Mismatch<2)
ZNF430.Intermediate.sites = subset(ZNF430.full.sites, between(Mismatch, 2, 4))
ZNF430.Weak.sites         = subset(ZNF430.full.sites, Mismatch>4)

ranges = GRangesList(ZNF430.Strong.sites, ZNF430.Intermediate.sites, ZNF430.Weak.sites)
names(ranges) = c("Strong ZNF430 sites", "Intermediate ZNF430 sites", "Weak ZNF430 sites")


soGGi::regionPlot(bamFile = "../Chromatin signals/Human/H3K9me3/ENCFF073GYV.FC.Trophoblast(20weeks).bigWig",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast tissue (20 weeks)",
                  distanceAround = 4000, format = "bigwig") -> H3K9me3.Trophoblast.20.signals

soGGi::regionPlot(bamFile = "../Chromatin signals/Human/H3K9me3/ENCFF040IBX.FC.Trophoblast(40weeks).bigWig",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast tissue (40 weeks)",
                  distanceAround = 4000, format = "bigwig") -> H3K9me3.Trophoblast.40.signals

soGGi::regionPlot(bamFile = "../Chromatin signals/Human/H3K9me3/ENCFF509OEW.FC.Hepatocyte.bigWig",
                  testRanges = ZNF430.full.sites,
                  samplename = "Hepatocytes", 
                  distanceAround = 4000, format = "bigwig") -> H3K9me3.Hepatocyte.signals

c(H3K9me3.Trophoblast.20.signals,
  H3K9me3.Trophoblast.40.signals,
  H3K9me3.Hepatocyte.signals) %>%
  soGGi::plotRegion(groupBy = "Sample",
                  gts = ranges,
                  colourBy = "Group",
                  freeScale = TRUE) +
  theme_bw() + ylab("Fold change over control") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0.15), limits = c(0.6, 2.45))+
  theme(panel.grid.minor = element_blank(), axis.title.x = element_blank())
#ggsave("H3K9me3 signal in placenta.svg", height = 3.3, width = 12)
```


## Aggregate signals around each class of sites in human trophoblast stem cells

This dataset is publicly available from [NCBI GEO database](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE200760)
```{r message=FALSE, warning=FALSE, fig.height=6.6, fig.width=12}
require(soGGi)

soGGi::regionPlot(bamFile = "Branco Lab Data/GSM6043306_hTSC_CutRun_H3K9me3.bw",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast stem cells (H3K9me3, Cut&Run)",
                  distanceAround = 4000, format = "bigwig") -> H3K9me3.hTSC.signals

soGGi::regionPlot(bamFile = "Branco Lab Data/GSM6043305_hTSC_CutRun_H3K27me3.bw",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast stem cells (H3K27me3, Cut&Run)",
                  distanceAround = 4000, format = "bigwig") -> H3K27me3.hTSC.signals

soGGi::regionPlot(bamFile = "Branco Lab Data/GSM6043307_hTSC_CutRun_IgG.bw",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast stem cells (IgG, Cut&Run)",
                  distanceAround = 4000, format = "bigwig") -> IgG.hTSC.signals

soGGi::regionPlot(bamFile = "Branco Lab Data/GSM6043300_hTSC_ChIP_H3K27ac.bw",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast stem cells (H3K27ac, ChIP)",
                  distanceAround = 4000, format = "bigwig") -> H3K27ac.hTSC.signals

soGGi::regionPlot(bamFile = "Branco Lab Data/GSM6043302_hTSC_ChIP_H3K4me3.bw",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast stem cells (H3K4me3, ChIP)",
                  distanceAround = 4000, format = "bigwig") -> H3K4me3.hTSC.signals

soGGi::regionPlot(bamFile = "Branco Lab Data/GSM6043303_hTSC_ChIP_Input.bw",
                  testRanges = ZNF430.full.sites, 
                  samplename = "Trophoblast stem cells (Input, ChIP)",
                  distanceAround = 4000, format = "bigwig") -> Input.hTSC.signals

require(ggplot2)
c(H3K27ac.hTSC.signals,
  H3K4me3.hTSC.signals,
  Input.hTSC.signals,
  H3K9me3.hTSC.signals,
  H3K27me3.hTSC.signals,
  IgG.hTSC.signals) %>%
  soGGi::plotRegion(groupBy = "Sample",
                  gts = ranges,
                  colourBy = "Group",
                  freeScale = FALSE) +
  theme_bw() + ylab("CPM") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 0.35))+
  theme(panel.grid.minor = element_blank(), axis.title.x = element_blank())

#ggsave("H3K9me3 signal in trophoblast stem cells.svg", height = 6.6, width = 12)
```

